---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# microEDA

<!-- badges: start -->
[![GitHub Release](https://img.shields.io/github/release/jrotzetter/microEDA?include_prereleases=&sort=semver&color=blue)](https://github.com/jrotzetter/microEDA/releases/ "View releases")
[![License](https://img.shields.io/badge/License-MIT-blue)](#license "View license summary")
[![Issues - microEDA](https://img.shields.io/github/issues/jrotzetter/microEDA)](https://github.com/jrotzetter/microEDA/issues "View open issues")
[![Made with R](https://img.shields.io/badge/R-4.5.3-blue?logo=r&logoColor=white)](https://cran.r-project.org/ "Go to CRAN homepage")
[![Made with R](https://img.shields.io/badge/RStudio-2026.01.1_Build_403-blue?logo=rstudio&logoColor=white)](https://posit.co/products/open-source/rstudio/ "Go to RSTUDIO IDE homepage")
<!-- badges: end -->

## Overview

R package for microbiome exploratory data analysis. Currently under active development.

## Installation

You can install the development version of microEDA from [GitHub](https://github.com/jrotzetter/microEDA) with:

```{r install_devtools, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jrotzetter/microEDA", build_vignettes = TRUE)
```

Alternatively you can also use the `pak` package:
```{r install_pak, eval = FALSE}
# install.packages("pak")
pak::pak("jrotzetter/microEDA")
```

Or the `remotes` package:
```{r install_remotes, eval=FALSE}
# install.packages("remotes")
remotes::install_github("jrotzetter/microEDA", build_vignettes = TRUE)
```
