---
title: "Introduction to microEDA"
output:
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Introduction to microEDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, # Global width
  fig.height = 5, # Global height
  fig.align = "center" # Optional: centers plots
)
```

`microEDA` simplifies exploratory data analysis (EDA) for microbiome datasets. Whether you're working with MetaPhlAn outputs or `phyloseq` objects, `microEDA` provides a unified interface for visualization, basic quality control, and taxonomic filtering. This vignette demonstrates how to import data, create `microEDA` objects, and use core utility and plotting functions.   

All functions work on both `microEDA` and `phyloseq` objects.

# Import MetaPhlAn Profiles

Since the phyloseq package does not natively support MetaPhlAn profiles, `microEDA` provides custom functions to load these files and merge them into a single `metaphlanProfile` object.
<br>
<br>
__A note on Database Version Compatibility:__
MetaPhlAn output files include a header line specifying the database version used for classification (e.g., `#mpa_vJun23_CHOCOPhlAnSGB_202307`). The `load_metaphlan()` function automatically extracts and stores this metadata.

When merging profiles with `join_mpa_profiles()`, the following rules apply:

* **Version Mismatch:** Profiles with different database versions cannot be merged and will throw an error.
* **Invalid Inputs:** Any input object that is not of class `metaphlanProfile` is ignored with a warning.

These checks ensure that only comparable profiles are merged, preventing errors arising from incompatible clade definitions or abundance normalizations.

```{r metaphlan}
library(microEDA)

# Load individual MetaPhlAn profiles
# load_metaphlan() automatically extracts the database version from the file header
mpa_merged <- load_metaphlan(system.file("extdata", "merged_abundance_table.txt", package = "microEDA"))

mpa_single <- load_metaphlan(system.file("extdata", "single_sample_profile.txt", package = "microEDA"))

# Merge profiles into a single metaphlanProfile object
# join_mpa_profiles() errors if database versions differ and warns on invalid inputs
mpa <- join_mpa_profiles(mpa_merged, mpa_single)
```

# Create microEDA-class Objects

microEDA objects can be easily created for objects of class `metaphlanProfile` or `phyloseq`:

```{r}
data(GlobalPatterns, package = "phyloseq")

me <- microEDA(GlobalPatterns)

mpa <- microEDA(mpa)
```

In turn, a `microEDA` or `metaphlanProfile` object can be converted to `phyloseq` with `to_phyloseq()`:

```{r}
ps <- to_phyloseq(me)

identical(GlobalPatterns, ps)
```

# Utility Functions

QIIME-style two-underscore prefixes (e.g., "k__", "p__") can be...
```{r}
# ...added to values in a taxonomyTable...
me <- add_taxonomy_prefix(me)
head(phyloseq::tax_table(me)[, c("Kingdom", "Phylum", "Order")])


# ... and just as easily be removed.
me <- trim_taxonomy_prefix(me)
head(phyloseq::tax_table(me)[, c("Kingdom", "Phylum", "Order")])
```

Inconsistencies (from, e.g., different database versions) in taxonomic annotations within the dataset at a specified rank can be identified with:

```{r}
check_taxonomic_consistency(me, tax_rank = "Genus", detailed_report = TRUE)
```

Taxa from a `microEDA` or `phyloseq` object can be filtered based on minimum abundance and prevalence thresholds. Filtering can be applied globally or within groups defined by a metadata variable. Optionally keeps filtered-out features in the info slot of `microEDA` objects.

```{r}
filtered_me <- filter_features(me, min_abundance = 100, min_prevalence = 0.1)
```

The dataset can also be collapsed at specific ranks by aggregating features of the same taxonomy into higher-level taxonomic groups.

```{r}
me_phylum <- agglomerate_taxa(me, tax_rank = "Phylum")
me_phylum
```

## Visualization Functions

```{r}
plot_taxa_barchart(me, "p", as_relative = FALSE)
```

```{r}
me_subset <- phyloseq::subset_samples(me, SampleType %in% c("Soil", "Freshwater", "Ocean", "Mock"))

plot_taxa_heatmap(me_subset, "Species", group_var = "SampleType", prevalence_display = "fraction", rm_missing = TRUE, as_relative = TRUE)
```

```{r}
plot_taxa_sankey(GlobalPatterns, tax_rank = "Order", samples = "Even3", as_relative = TRUE)
```

```{r}
plot_taxa_upset(GlobalPatterns, "Phylum", "SampleType")
```

To retrieve the specific names of taxa shared between or unique to groups, use `get_taxa_overlaps()` instead of `plot_taxa_upset()`:

```{r}
# Calculate overlaps for Phyla across SampleTypes
overlaps <- get_taxa_overlaps(GlobalPatterns, "Phylum", "SampleType",
  min_abundance = 100, min_prevalence = 3
)

# View taxa specific to "Soil"
overlaps[["Soil-specific"]]

# View taxa shared by all groups
overlaps[["all-intersection"]]
```
